---
title: "PA1_template"
author: "Charles Breeden"
date: "December 10, 2017"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

##Introduction
This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

##Loading and preprocessing the data
First, the data for this assignment was downloaded from the course web site, processed to filter out N/A values and summarized by total steps per day.

```{r, echo = TRUE, warning = FALSE}
library(dplyr)
library(ggplot2)
library(xtable)
library(lubridate)

path <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
temp <- tempfile()
download.file(path, temp)
activity <- read.csv(unz(temp, "activity.csv"))
unlink(temp)
```

##Calculating mean total number of steps taken per day
This section calculates the total number of steps taken per day. The following plot shows the total steps taken each day and the table summarizes the average and median total steps per day.

```{r, results = 'asis', echo = TRUE}
graph <- activity[!is.na(activity$steps), ]
graph <- graph %>% group_by(date) %>% summarize(tot_steps = sum(steps))

par(mfrow = c(1, 1))

g <- ggplot(data = graph, aes(x = tot_steps)) 
g <- g + labs(x = "Total Steps", y = "Days", title = "Histogram of Total Daily Steps")
g <- g + geom_histogram(fill = "darkred", color = "lightgray") 
g + theme(plot.title = element_text(hjust = 0.5))

# Calculate mean and median
t <- graph %>% summarize(mean_steps = mean(tot_steps), median_steps = median(tot_steps))
colnames(t) <- c("Mean Steps", "Median Steps")
print(xtable(t), type = "html", html.table.attributes = "border=1")
```

##Calculate the average daily activity pattern
This section plots for each 5-minute interval the average number of steps for all days in the dataset and summarizes the maximum average number of steps, along with the corresponding interval associated with this maximum.

```{r, results = 'asis', echo = TRUE}
graph <- activity %>% group_by(interval) %>% summarize(avg_steps = mean(steps, na.rm = TRUE))

g <- ggplot(data = graph, aes(x = interval, y = avg_steps)) 
g <- g + geom_line(color = "blue")
g <- g + labs(x = "Interval", y = "Average Steps per Interval", title = "Average Steps per Interval")
g + theme(plot.title = element_text(hjust = 0.5))

t <- graph[graph$avg_steps == max(graph$avg_steps), ]
colnames(t) <- c("Interval", "Max Steps")
print(xtable(t), type = "html", html.table.attributes = "border=1")
```

##Impute missing values
This section calculates and reports on the total number of missing values in the dataset, uses the average number of steps per interval to impute these missing values, creates a new dataset that is equal to the original dataset but with the missing data filled in, makes a histogram of the total number of steps taken each day and calculates and reports the mean and median total number of steps taken per day.

```{r, results = 'asis', echo = TRUE}
# Calculate and report the total number of missing values in the dataset 
r <- nrow(activity[is.na(activity$steps), ])

# Devise a strategy for filling in all of the missing values in the dataset
# Create a new dataset that is equal to the original dataset but with the missing data filled in
new_act <- activity
new_act$missing = is.na(new_act$steps)
new_act[is.na(new_act$steps), ]$steps = graph[graph$interval %in% new_act$interval, ]$avg_steps

# Make a histogram of the total number of steps taken each day 
graph <- new_act
graph <- graph %>% group_by(date) %>% summarize(tot_steps = sum(steps))

g <- ggplot(data = graph, aes(x = tot_steps)) 
g <- g + geom_histogram(fill = "darkred", color = "lightgray") 
g <- g + labs(x = "Total Steps", y = "Days", title = "Histogram of Total Daily Steps")
g + theme(plot.title = element_text(hjust = 0.5))

# Calculate and report the mean and median total number of steps taken per day
t <- graph %>% summarize(mean_steps = mean(tot_steps), median_steps = median(tot_steps))
colnames(t) <- c("Mean Steps", "Median Steps")
print(xtable(t), type = "html", html.table.attributes = "border=1")
```

##Determine differences in activity patterns between weekdays and weekends
This section creates a new factor variable in the data set with two levels - "weekday" and "weekend" and makes a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r, echo = TRUE}
# Create a new factor variable in the dataset with two levels - "weekday" and "weekend" 
# indicating whether a given date is a weekday or weekend day

new_act$day = weekdays(as.Date(as.character(new_act$date)))
new_act$daytype = "weekday"
new_act[wday(new_act$date, week_start = 1) > 5, ]$daytype = "weekend"
new_act$daytype <- as.factor(new_act$daytype)

graph <- new_act
graph <- graph %>% group_by(interval, daytype) %>% summarize(avg_steps = mean(steps))

g <- ggplot(data = graph, aes(x = interval, y = avg_steps)) 
g <- g + facet_grid(facets = .~daytype) + geom_line(color = "blue") 
g <- g + labs(x = "Interval", y = "Average Steps per Interval", 
     title = "Average Steps per Interval, Weekday vs. Weekend")
g + theme(plot.title = element_text(hjust = 0.5))
```